<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{An Introduction to MANOVA.RM}
-->


## Introduction

This vignette documents the use of the `MANOVA.RM` package for the analysis of semi-parametric repeated measures designs and multivariate data. 
The package consists of two important functions, which will be explained in detail below. Both functions calculate the Wald-type statistic 
(WTS) and the ANOVA-type statistic (ATS) based on means. Furthermore, different resampling approaches are provided in order to improve the
small sample behavior of the WTS. These test statistics can be used for arbitrary semi-parametric designs, even with unequal covariance matrices among groups
and small sample sizes.

## The `RM` function

The `RM` function calculates the above mentioned test statistics in a repeated measures design with an arbitrary number of crossed whole-plot and 
sub-plot factors.
The resampling methods provided are a permutation procedure, a parametric bootstrap approach and a
wild bootstrap using Rademacher weights. The wild bootstrap is also implemented for the ATS.


### Data Example 1 (One whole-plot and two sub-plot factors)

For illustration purposes, we consider the data set `o2cons`, which is included in `MANOVA.RM`. 

```{r}
library(MANOVA.RM)
data(o2cons)
```

The data set contains measurements of the oxygen consumption of leukocytes in the presence and absence of inactivated staphylococci 
at three consecutive time points. 
Due to the study design, both time and staphylococci are sub-plot factors while the treatment (Verum vs. Placebo) is a whole-plot
factor.

```{r}
head(o2cons)
```

We will now analyze this data using the `RM` function.
The `RM` function takes as arguments:
* `formula`: A formula consisting of the outcome variable on the left hand side of a \~ operator and the factor
variables of interest on the right hand side. An interaction term must be specified.
* `data`: A data.frame, list or environment containing the variables in `formula`.
* `subject`: The column name of the subjects variable in the data frame.
* `no.subf`: The number of sub-plot factors, default is 1.
* `iter`: The number of iterations for the resampling approach. Default value is 10000.
* `alpha`: The significance level, default is 0.05.
* `resampling`: The resampling method, one of 'Perm', 'paramBS' or 'WildBS'. Default is
set to 'Perm'.


```{r}
set.seed(1234)
model1 <- RM(O2 ~ Group * Staphylococci * Time, data = o2cons, 
             subject = "Subject", no.subf = 2, iter = 10000, resampling = "Perm", CPU = 1)
summary(model1)
```

The output consists of four parts: `model1$Descriptive` gives an overview of the descriptive statistics: The number of observations, 
mean and confidence intervals (based on quantiles of the t-distribution) are displayed for each factor level combination.
`model1$WTS` contains the results for the Wald-type test: The test statistic, degree of freedom and p-values based on the asymptotic $\chi^2$ distribution
are displayed. Note that the $\chi^2$ approximation is very liberal for small sample sizes.
`model1$ATS` contains the corresponding results based on the ATS. 
This test statistic tends to rather
conservative decisions in the case of small sample sizes and is even asymptotically only an approximation, thus not providing an asymptotic level $\alpha$ test.
Finally, `model1$resampling` contains the p-values based on 
the chosen resampling approach. For the ATS, only the wild bootstrap procedure can be applied.
Due to the above mentioned issues for small sample sizes, the resampling procedure is recommended for such situations.

### Data Example 2 (Two sub-plot and two whole-plot factors)

We consider the data set `EEG` from the `MANOVA.RM` package: At the Department of Neurology, University Clinic of Salzburg, 160 patients were diagnosed
with either AD, MCI, or SCC, based on neuropsychological diagnostics (Bathke et al.(2015), preprint). This data set contains z-scores for 
brain rate and Hjorth complexity,
each measured at frontal, temporal and central electrode positions and averaged across hemispheres. In addition to standardization, complexity
values were multiplied by -1 in order to make them more easily comparable to brain rate
values: For brain rate we know that the values decrease with age and pathology, while
Hjorth complexity values are known to increase with age and pathology.
The three between-subjects factors considered were sex (men vs. women), diagnosis (AD
vs. MCI vs. SCC), and age ($< 70$ vs. $>= 70$ years). Additionally, the within-subjects factors region (frontal, temporal, central) and 
feature (brain rate, complexity) structure the response vector.

```{r}
data(EEG)
set.seed(987)
EEG_model <- RM(resp ~ sex * diagnosis * feature * region, 
                     data = EEG, subject = "id", no.subf = 2, resampling = "WildBS",
                     iter = 1000,  alpha = 0.01, CPU = 1)
summary(EEG_model)
```

We find significant effects at level $\alpha = 0.01$ of the whole-plot factors sex and diagnosis, while none of the sub-plot factors or interactions become significant.

### Plotting

The `RM()` function is equipped with a plotting option, displaying the calculated means along with $(1-\alpha)$ confidence intervals.
The `plot` function takes an `RM` object as an argument. In addition, the factor of interest may be specified. If this argument is 
omitted in a two- or higher-way layout, the user is asked to specify the factor for plotting. Furthermore, additional graphical parameters
can be used to customize the plots. The optional argument `legendpos` specifies the position of the legend in higher-way layouts.

```{r}
plot(EEG_model, factor = "sex", main = "Effect of sex on EEG values")
plot(EEG_model, factor = "sex:diagnosis", legendpos = "topleft", col = c(4, 2))
plot(EEG_model, factor = "sex:diagnosis:feature", legendpos = "center")
```



## The `MANOVA` function

The `MANOVA` function calculates the above mentioned test statistics for multivariate data in a design with crossed or nested factors.
The resampling methods provided are a parametric bootstrap approach and a
wild bootstrap using Rademacher weights. The wild bootstrap is also implemented for the ATS, while the parametric approach works only for the WTS,
see Konietschke et al. (2015) for details. 
Note that only balanced nested designs (i.e., the same number of factor levels $b$ for each level of the factor $A$) with up to three factors are 
implemented. Designs involving both crossed and nested factors are not implemented.


### Data Example MANOVA (two crossed factors)

We again consider the data set `EEG` from the `MANOVA.RM` package, but now we ignore the sub-plot factor structure. Therefore, we are
now in a multivariate setting with 6 measurements per patient and three crossed factors sex, age and diagnosis. Due to the small number of subjects in
some groups (e.g., only 2 male patients aged $<$ 70 were diagnosed with AD) we restrict our analyses to two factors at a time.
The analysis of this example is shown below.

The `MANOVA` function takes as arguments:
* `formula`: A formula consisting of the outcome variable on the left hand side of a \~ operator and the factor
variables of interest on the right hand side. An interaction term must be specified.
* `data`: A data.frame, list or environment containing the variables in `formula`.
* `subject`: The column name of the subjects variable in the data frame.
* `iter`: The number of iterations for the resampling procedure. Default value is 10000.
* `alpha`: The significance level, default is 0.05.
* `resampling`: The resampling method, one of 'paramBS' and 'WildBS'. Default is
set to 'paramBS'.


```{r}
data(EEG)
set.seed(987)
EEG_MANOVA <- MANOVA(resp ~ sex * diagnosis, 
                     data = EEG, subject = "id", resampling = "paramBS", 
                     iter = 1000,  alpha = 0.01, CPU = 1)
summary(EEG_MANOVA)
```



## optional GUI

The `MANOVA.RM` package is equipped with an optional graphical user interface, which is based on `RGtk2`. The GUI may be started in `R` (if `RGtk2` is installed) using the
command `GUI.RM()` and `GUI.MANOVA()` for repeated measures designs and multivariate data, respectively. 

```{r}
GUI.MANOVA()
```

The user can specify the data location
(either directly or via the "load data" button), the formula, the number of iterations for the resampling approach and
the significance level. Furthermore, one needs to specify the number of sub-plot factors (for the repeated measures design only),
the 'subject' variable in the 
data frame and the resampling method.
Additionally, one can specify whether or not headers are
included in the data file, and which separator (e.g., ',' for *.csv files) and character symbols are used for decimals
in the data file. 
The GUI for `RM` also provides a plotting option, which generates a new window
for specifying the factors to be plotted (in higher-way layouts) along with a few plotting
parameters.
